---
title: "class05: data viz with ggplot"
author: "Ryan Bench (PID:A69038034)"
format: gfm
---

Today we are playing with plotting and graphics in R.

There are lots of ways to make cool figures in R.
There is "base" R graphics (`plot()`, `hist()`, `boxplot()` etc.)

There is also add-on packages, like **ggplot**.

```{r}
head(cars, 3)
```

Let's plot this with "base" R: 
control+alt+i is the shortcut for inserting code chunks in r
```{r}
plot(cars)
```

```{r}
head(mtcars)
```
Let's plot `mpg` vs `disp`

```{r}
plot(mtcars$mpg, mtcars$disp, pch = 15)
```

```{r}
hist(mtcars$mpg)
```
## GGPLOT

The main function in the ggplot2 package is **ggplot()**.  First I need to install the **ggplot2** package. I can install any package with the function `install.packages("ggplot2")`. This is a one time install then it's on my computer. Use the terminal to install the package.

> **N.B.** I never want to run `install.packages()` in my quarto source document!!!

Every ggplot needs 3 things, data source, aesthetic, and geometry
```{r}
library(ggplot2)
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()
```

```{r}
ggplot(cars) + 
  aes(speed) +
  geom_histogram()
```

Every GGPLOT needs at least three things:
- The **data** (given with `ggplot(cars)`)
- The **aesthetic** mapping (given with `aes()`)
- The **geom** (given by `geom_point()`)

> For simple canned graphs "base" R is nearly always faster

### Adding more layers

Let's add a line and a title, subtitle, and caption as well as custom axis labels.

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()+
  geom_line()
```

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()+
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Silly Plot", 
       x = "Speed (MPH)", 
       y = "Distance (ft)", 
       caption = "when is tea time anyway??") + 
  theme_bw()
```

##Plot some expression data

Read data file from online URL

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this wee dataset?

There are `r nrow(genes)` in this dataset.

> Q2. How many "up" regulated genes are there?

```{r}
sum(genes$State == "up")
```

There are `r sum(genes$State == "up")`

```{r}
table(genes$State)
```

```{r}
library(ggrepel)
ggplot(genes) + 
  aes(x=Condition1, 
      y=Condition2, 
      col = State, 
      label = Gene) +
  geom_point() +
  scale_color_manual(values=c("blue", "gray", "red")) +
  geom_text_repel(max.overlaps = 60) + 
  theme_bw()
  labs(title = "Gene Expression Changes Upon Drug Treatment", 
       x = "Control(No Drug)", 
       y = "Drug Treatment")
  
```

##Going further

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```

```{r}
tail(gapminder)
```

```{r}
ggplot(gapminder) + 
  aes(y=lifeExp, 
      x=gdpPercap,
      color = continent) +
  geom_point(alpha=0.2) +
  facet_wrap(~continent) +
  theme_bw()
```

```{r}
ggplot(gapminder) + 
  aes(y=lifeExp, 
      x=year,
      color = continent) +
  geom_point() +
  facet_wrap(~continent) +
  theme_bw() +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
# Setup some example plots 
library(patchwork)
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))

# Use patchwork to combine them here:
(p1 | p2 | p3) /
      p4
```

