---
title: "class 06 R functions"
author: "Ryan Bench (PID:A69038034)"
format: pdf
toc: true
---

All functions in R have at least 3 things

- A **name**, we pick this and use it to call our function,
- Input **arguments** (there can be multiple)
- The **body** lines of R code that do the work


## Our first (silly) function

Write a function to house some numbers

```{r}
add <- function(x, y=1) {
  x + y
}
```

If we have (x, y=1), if y is not defined it just defaults to 1
Now we can call the function:

```{r}
add(c(10, 10), 100)
```


## A second function

We are writing a function that will generate random nucleotide sequences of a user specified length

The `sample` function can be helpful here

```{r}
sample(c("A", "C", "T", "G"), size=50, replace = TRUE)
```

I want a 1 element long character vector that looks like this "CACAGC" not "C" "A" "C" "A" "G" "C"

```{r}
v <- sample(c("A", "C", "T", "G"), size=50, replace = TRUE)
paste(v, collapse = "")
```

```{r}
generate_DNA <- function(size) {
  v <- sample(c("A", "C", "T", "G"), size = size, replace = TRUE)
  paste(v, collapse = "")
}
```

Test it:

```{r}
generate_DNA(100)
```

```{r}
fasta <- FALSE
if(fasta) {
  cat("HELLO You!")
} else {
  cat("No you dont!")
}
```

Add the ability to return a multi-element vector or a single element fasta like vector.

```{r}
generate_fasta <- function(size = 50, fasta = TRUE) {
  v <- sample(c("A", "C", "T", "G"), size = size, replace = TRUE)
  s <- paste(v, collapse = "")
  
  if(fasta) {
    return(s)
  } else{
    return(v)
  }
} 
```

```{r}
generate_fasta(60, fasta = TRUE)
```

```{r}
generate_fasta(60, fasta = FALSE)
```

## A protein generating function 

```{r}
generate_protein <- function(size = 50, fasta = TRUE) {
  aa <- c("A", "R", "N", "D", "C", "Q", "E", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
  v <- sample(aa, size = size, replace = TRUE)
  s <- paste(v, collapse = "")
  
  if(fasta) {
    return(s)
  } else {
    return(v)
  }
}
```

```{r}
generate_protein(60)
```

Use our new "generate_protein()" function to make random protein sequences of length 6-12 (i.e. one length 6, one length 7, etc up to length 12)

This can be done via "brute force" 

```{r}
generate_protein(6)
generate_protein(7)
generate_protein(8)
generate_protein(9)
generate_protein(10)
generate_protein(11)
generate_protein(12)
```

A second way to do this is to use the `for()` loop:

```{r}
lengths <- 6:12
lengths
```

```{r}
for(i in lengths) {
  cat(">", i, "\n", sep="")
 aa <- generate_protein(i)
 cat(aa)
  cat("\n")
}
```

```{r}
sapply(6:12, generate_protein)
```

